# Writing with Quarto {#sec-guide}

This chapter is a worked example. Open `guide.qmd` next to the PDF to see how
each element is written, copy what you need into your own chapters, then
delete this one and remove it from `chapters:` in `_quarto.yml`. Writing text
and code in one document goes back to what @knuth84 called literate
programming: the analysis and its description can no longer drift apart.

The example asks an old question in empirical economics: do countries with
more dependants save less? Under the life-cycle hypothesis
[@modigliani1966], people save during their working years and run their
savings down in retirement, so countries with more young and old dependants
should save a smaller share of their income. The data are R's built-in
`LifeCycleSavings` set, 50 countries averaged over 1960 to 1970, which
@belsley1980 took from a bachelor's thesis written at MIT.^[Footnotes like
this one are written inline, right where they belong, and numbered
automatically.]

Load the packages and prepare the data once, at the top of the chapter:

```{r}
#| label: setup
#| message: false

library(ggplot2)
library(modelsummary)

# Income in thousands, so its coefficient is readable
savings <- LifeCycleSavings
savings$dpi_k <- savings$dpi / 1000
```

## Citations {#sec-citations}

References live in `references.bib`, a plain-text BibTeX file that a
reference manager such as Zotero (with the Better BibTeX add-on) can keep up
to date for you. Every entry has a key, and you cite it by writing the key
after an `@`:

| You write                  | You get                  |
|----------------------------|--------------------------|
| `@modigliani1966 argues`   | @modigliani1966 argues   |
| `the data [@belsley1980]`  | the data [@belsley1980]  |
| `[see @modigliani1966]`    | [see @modigliani1966]    |
| `[@knuth84; @r2024]`       | [@knuth84; @r2024]       |

Add a page number with `[@modigliani1966, p. 12]`. The list of references at
the end holds every source you cite and nothing else, in the Chicago
author-date style unless you choose another with `csl:` in `_quarto.yml`.
Cite your software too: `citation()` and `citation("modelsummary")` in R
print ready-made entries.

## Equations {#sec-equations}

Equations use LaTeX notation between double dollar signs. A label after the
closing `$$` numbers the equation:

$$
\text{sr}_i = \beta_0 + \beta_1\,\text{pop15}_i + \beta_2\,\text{pop75}_i
  + \beta_3\,\text{dpi}_i + \beta_4\,\text{ddpi}_i + \varepsilon_i
$$ {#eq-lifecycle}

Here $\text{sr}_i$ is the savings ratio of country $i$, aggregate personal
saving as a percentage of disposable income; pop15 and pop75 are the shares
of the population under 15 and over 75; dpi is real disposable income per
head and ddpi its growth rate. Inline maths such as $\beta_1 < 0$ goes
between single dollar signs.

## Figures {#sec-figures}

Figures come from code chunks. The options at the top of a chunk, the lines
starting with `#|`, give the figure a label that begins with `fig-` and a
caption, and @fig-savings is then numbered and can be referenced anywhere.

```{r}
#| label: fig-savings
#| fig-cap: "Savings ratio and the share of the population under 15, 50 countries, averages for 1960 to 1970"
#| fig-width: 6
#| fig-height: 3.5

ggplot(savings, aes(x = pop15, y = sr)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x,
              colour = "black") +
  labs(x = "Population under 15 (%)",
       y = "Savings ratio (%)") +
  theme_minimal()
```

Code is shown here so that you can see how each result is made. To hide it
in your thesis, add `#| echo: false` to a chunk, or `echo: false` under
`execute:` in `_quarto.yml` for all chunks at once.

## Tables {#sec-tables}

Tables work the same way, with a label that begins with `tbl-`. The
modelsummary package [@arelbundock2022] turns fitted models into regression
tables that look right in both the PDF and the web version:

```{r}
#| label: tbl-savings
#| tbl-cap: "The savings ratio and the age structure of the population, OLS estimates of @eq-lifecycle"
#| tbl-pos: "H"

models <- list(
  "(1)" = lm(sr ~ pop15 + pop75, data = savings),
  "(2)" = lm(sr ~ pop15 + pop75 + dpi_k + ddpi,
             data = savings)
)

modelsummary(
  models,
  stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
  coef_map = c(
    pop15 = "Population under 15 (%)",
    pop75 = "Population over 75 (%)",
    dpi_k = "Disposable income per head (thousands)",
    ddpi = "Growth of disposable income (%)",
    "(Intercept)" = "Constant"
  ),
  gof_map = c("nobs", "r.squared")
)
```

```{r}
#| include: false

b_pop15 <- coef(models[["(2)"]])[["pop15"]]
```

In column (2) of @tbl-savings, a share of under-15s one percentage point
higher goes with a savings ratio `r round(abs(b_pop15), 2)` points lower.
That number is not typed in: a short piece of R code inside the sentence
computes it, so it changes whenever the data or the model do. Standard errors
are in parentheses; `vcov = "HC1"` in `modelsummary()` switches to
heteroskedasticity-robust ones (this needs the sandwich package). The
option `#| tbl-pos: "H"` keeps the table where it is written; without it,
LaTeX may move a table to the top or bottom of a page. Descriptive statistics for all five variables are in @tbl-descriptives in
the appendix.

## Cross-references {#sec-crossrefs}

Anything with a label can be referenced with `@` and its label: writing
`@fig-savings`, `@tbl-savings`, `@eq-lifecycle` and `@sec-citations` gives
@fig-savings, @tbl-savings, @eq-lifecycle and @sec-citations. Section labels
such as `{#sec-citations}` go after a heading. The numbers update when you
add, move or delete figures, tables and sections, so never type them by hand.

## Workflow {#sec-workflow}

A few habits keep a thesis reproducible from the first draft to the final
PDF:

- **Render often.** `quarto preview` shows the web version and refreshes it
  whenever you save; `quarto render`, or Render Book in RStudio's Build pane,
  builds the PDF in `_book/`. Problems are easiest to fix while they are
  small.
- **Keep data and code in the project.** Put raw data in a `data/` folder,
  read it with relative paths such as `read.csv("data/survey.csv")`, and
  never edit it by hand. Each chapter runs in its own R session, so load the
  packages and data it needs at its top, as this one does.
- **Let the computer remember.** `freeze: auto` in `_quarto.yml` re-runs a
  chapter's code only when that chapter changes. The template does not pin
  package versions; if your results must reproduce exactly years from now,
  [renv](https://rstudio.github.io/renv/) or
  [groundhog](https://groundhogr.com/) can do that for R packages, and
  [Docker](https://www.docker.com/) for the whole computing environment.
- **Use Git.** Commit whenever you finish a piece of work, with a message
  that says what changed, and push to GitHub, so that every version of your
  thesis is backed up.
